---
title: "BIOL 220 Hypothesis Testing"
output: binb::metropolis
header-includes:
  - \usepackage{array}
  - \usepackage{xcolor}
  - \usepackage[export]{adjustbox}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
                      eval = TRUE)
library(cowplot)
library(ggplot2)
library(dplyr)
library(viridis)
```



## Learning Objectives

Last lecture:

1. Identify an appropriate \alert{scientific hypothesis} that addresses a biological question
3. Identify an appropriate \alert{Null Hypothesis} that tests a scientific hypothesis
4. Identify an \alert{Alternative Hypothesis}

This lecture: 

4. Understand the practice of \alert{Null Hypothesis Significance Testing} (NHST)
5. Explain the meaning of a \alert{$P$-value}
7. Explain the concepts of \alert{false positives} and \alert{false negatives} and why they happen


## Identify a \alert{Null Hypothesis}


> A \alert{null hypothesis} is a specific statement about a population parameter made for the purposes of argument. A good \alert{null hypothesis} is a statement that would be interesting to reject


## A null hypothesis must be specific and should be "interesting" to reject {.t}

**For example:** 

There are 0 trees in Hawaiʻi

This is specific, but trivially easy and uninteresting to reject

## A null hypothesis must be specific and should be "interesting" to reject {.t}


**For example:**

The growth rate of native trees in Hawaiʻi are the same as their con-generic relatives from continents

This is specific, *and* interesting to reject


## Identify an \alert{Alternative Hypothesis}

> The \alert{alternative hypothesis} includes all other feasible values for the population parameter besides the value stated in the null hypothesis

## An alternative hypothesis is not specific

**For example**:

The growth rate of native trees in Hawaiʻi are *different* than their con-generic relatives from continents


## Naming conventions

$H_0$: pronounced "H-naught" or "H-zero" is the convention for the \alert{null hypothesis}

$H_\mathrm{A}$:is the convention for the \alert{alternative hypothesis}



## Identifying \alert{null} and \alert{alternative} hypothesis

**Example: noni**

::: columns

:::: column

```{r, fig.width=3, fig.height=4}
set.seed(1)
dat <- data.frame(sed = c(rnorm(100, -2), rnorm(100)), 
                  group = c(rep("10mg noni", 100), 
                            rep("placebo", 100)))
noni_dat <- dat

ggplot(dat, aes(x = sed, fill = group)) +
    geom_histogram(aes(y = after_stat(density)), bins = 20) +
    facet_grid(rows = vars(group)) +
    theme_cowplot() +
    theme(legend.position = "none") +
    xlab("Change in sed rate [mm/hr]\n(after - before)")

```

::::

:::: column

\metroset{block=fill}

\begin{alertblock}{Pause \& Think}

\begin{itemize}
\item{What is the \alert{null hypothesis}?}
\item{What is the \alert{alternative hypothesis}?}
\end{itemize}

\end{alertblock}

- $H_0$: The mean of *change in sedimentation rate* is the same across the two groups
- $H_A$: The mean is not the same

::::

:::


## \alert{Null Hypothesis Significance Testing}

What is \alert{Null Hypothesis Significance Testing}?

## \alert{Null Hypothesis Significance Testing}

What is \alert{Null Hypothesis Significance Testing}?  It asks the question:
\vspace{1em}

> What would be the probability of observing the sample estimate, or a more extreme estimate, if the \alert{null hypothesis} were true?

\vspace{2em}

> If that probability is small, we say "we reject the null hypothesis"

\vspace{2em}

> If that probability is not small, we say "we fail to reject the null hypothesis"


## \alert{Null Hypothesis Significance Testing}

What is \alert{Null Hypothesis Significance Testing}?  It asks the question:
\vspace{1em}

> What would be the probability of observing the sample estimate, or a more extreme estimate, if the \alert{null hypothesis} were true?

\vspace{2em}

The sample estimate of interest depends on the scientific and null hypotheses

## \alert{Null Hypothesis Significance Testing}


The sample estimate of interest depends on the scientific and null hypotheses

::: columns

:::: column

```{r, fig.width=3, fig.height=4}
p <- ggplot(dat, aes(x = sed, fill = group, alpha = 0.5)) +
    geom_histogram(aes(y = after_stat(density)), bins = 20) +
    facet_grid(rows = vars(group)) +
    theme_cowplot() +
    theme(legend.position = "none") +
    xlab("Change in sed rate [mm/hr]\n(after - before)") 

est <- group_by(dat, group) |>
    summarize(ybar = mean(sed))
ann <- data.frame(x = est$ybar, y = 0.5, group = est$group,
                  lab = c("bar(Y)['yes noni']", 
                          "bar(Y)['no noni']"))

p + geom_vline(est, mapping = aes(xintercept = ybar, color = group), size = 1.5) +
    geom_label(data = ann, 
               mapping = aes(x = x, y = y, label = lab), 
               parse = TRUE, hjust = "left", nudge_x = 0.1, 
               size = 5) +
    geom_segment(data = data.frame(group = "placebo"), 
                 mapping = aes(x = ann$x[1], xend = ann$x[2], 
                               y = 0.6, yend = 0.6), 
                 arrow = arrow(length = unit(0.1, "npc"), ends = "both"), 
                 size = 1.5) +
    geom_label(data = data.frame(group = "placebo"), 
               x = -2, y = 0.45, label = expression(bar(Y)["diff"]), 
               fill = "white", size = 5) +
    coord_cartesian(xlim = c(-4.5, 2.75), ylim = c(0, 0.55), clip = "off")
```

::::

:::: column

- $H_0$: The mean of *change in sedimentation rate* is the same across the two groups
- $H_A$: The mean is not the same
- **The estimate of interest:**

$\bar{Y}_\text{diff} = \bar{Y}_\text{yes noni} - \bar{Y}_\text{no noni}$

::::

:::


## \alert{Null Hypothesis Significance Testing}


> What would be the probability of observing the sample estimate, or a more extreme estimate, if the \alert{null hypothesis} were true?

\vspace{2em}

When doing \alert{Null Hypothesis Significance Testing}, the specific sample estimate we use is called a \alert{test statistic}

Here we decided the \alert{test statistic} is $\bar{Y}_\text{diff}$


## \alert{Null Hypothesis Significance Testing}


> What would be the probability of observing the sample estimate, or a more extreme estimate, if the \alert{null hypothesis} were true?

\vspace{2em}

How do we get the probability distribution of the \alert{test statistic}?


## \alert{Null Hypothesis Significance Testing}


> In many many **many** hypothetical repeated studies, where the \alert{null hypothesis} is TRUE, what would be the distribution of \alert{test statistics}?

\vspace{2em}

From that \alert{sampling distribution} of test statistics, we can calculate the probability of observing the test statistic, or a more extreme value



## Simulating a sampling distribution of the test statistic under the null hypothesis

\Large
\centering
"The Null Distribution"

## Simulating the null distribution

```{r null-dist, cache=TRUE, echo=TRUE}
# repeatedly shuffle the groups to make a null distribution
all_null <- replicate(1000, {
    noni_null <- noni_dat
    noni_null$group <- sample(noni_null$group)
    
    ybars_null <- group_by(noni_null, group)
    ybars_null <- summarize(ybars_null, ybar = mean(sed))
    test_stat_null <- diff(ybars_null$ybar)
    test_stat_null
})
```

## Simulating the null distribution

```{r echo = TRUE, fig.align='center', fig.width=3, fig.height=2}
null_dist <- data.frame(test_stat_null = all_null)

ggplot(null_dist, aes(x = test_stat_null)) + 
    geom_histogram(aes(y = after_stat(density)))
```

## Simulating the null distribution

```{r echo = TRUE, fig.align='center', fig.width=3, fig.height=2}
# add the observed test statistic
ybars <- group_by(noni_dat, group)
ybars <- summarize(ybars, ybar = mean(sed))
test_stat <- diff(ybars$ybar)
```

## Simulating the null distribution

```{r echo = TRUE, fig.align='center', fig.width=3, fig.height=2}
ggplot(null_dist, aes(x = test_stat_null)) + 
    geom_histogram(aes(y = after_stat(density))) +
    geom_vline(xintercept = test_stat, 
               color = "red", size = 1)
```


## Simulating the null distribution


> In many many **many** hypothetical repeated studies, where the \alert{null hypothesis} is TRUE, what would be the distribution of \alert{test statistics}?

\vspace{2em}

From that \alert{sampling distribution} of test statistics, we can calculate the probability of observing the test statistic, or a more extreme value

That probability is:

```{r echo = TRUE}
mean(null_dist$test_stat_null > test_stat | 
         null_dist$test_stat_null < -test_stat)
```

This probability has a special name: the \alert{p-value}


## \alert{Null Hypothesis Significance Testing}

Null Hypothesis Significance Testing asks the question:
\vspace{1em}

> What would be the probability (the \alert{p-value}) of observing the test statistic, or a more extreme value, if the \alert{null hypothesis} were true?

\vspace{2em}

> If the \alert{p-value} is small, we say "we reject the null hypothesis"

\vspace{2em}

> If the \alert{p-value} is not small, we say "we fail to reject the null hypothesis"

\large
\centering
How small is small enough?


## \alert{Null Hypothesis Significance Testing}


\begin{center}
\large
How small of a \alert{p-value} is small enough?
\end{center}

\normalsize

Answer: depends on the desired \alert{significance level}

- we write the significance level as $\alpha$
- in biology, $\alpha$ is often 0.05
- remember 95\% Confidence Intervals? Same thing as $\alpha = 0.05$

## Can we reject the null?

::: columns

:::: column

```{r, fig.width=3, fig.height=4}
p <- ggplot(dat, aes(x = sed, fill = group, alpha = 0.5)) +
    geom_histogram(aes(y = after_stat(density)), bins = 20) +
    facet_grid(rows = vars(group)) +
    theme_cowplot() +
    theme(legend.position = "none") +
    xlab("Change in sed rate [mm/hr]\n(after - before)") 

est <- group_by(dat, group) |>
    summarize(ybar = mean(sed))
ann <- data.frame(x = est$ybar, y = 0.5, group = est$group,
                  lab = c("bar(Y)['yes noni']", 
                          "bar(Y)['no noni']"))

p + geom_vline(est, mapping = aes(xintercept = ybar, color = group), size = 1.5) +
    geom_label(data = ann, 
               mapping = aes(x = x, y = y, label = lab), 
               parse = TRUE, hjust = "left", nudge_x = 0.1, 
               size = 5) +
    geom_segment(data = data.frame(group = "placebo"), 
                 mapping = aes(x = ann$x[1], xend = ann$x[2], 
                               y = 0.6, yend = 0.6), 
                 arrow = arrow(length = unit(0.1, "npc"), ends = "both"), 
                 size = 1.5) +
    geom_label(data = data.frame(group = "placebo"), 
               x = -2, y = 0.45, label = expression(bar(Y)["diff"]), 
               fill = "white", size = 5) +
    coord_cartesian(xlim = c(-4.5, 2.75), ylim = c(0, 0.55), clip = "off")
```

::::

:::: column

- $H_0$: The mean of *change in sedimentation rate* is the same across the two groups
- Test statistic: $\bar{Y}_\text{diff} = \bar{Y}_\text{yes noni} - \bar{Y}_\text{no noni}$
- Test statistic: $\bar{Y}_\text{diff} =$ `r round(test_stat, 2)`
- P-value: 0

\large
Yes!  At significance level $\alpha = 0.05$, we reject the null that noni has no impact on sedimentation rate

::::



:::


## \alert{Null Hypothesis Significance Testing}: the steps

1. State $H_0$ and $H_A$
2. Calculate the test statistic
3. Generate the null distribution
5. Calculate the p-value
6. Decide:
    - Reject $H_0$ if the p-value is less than or equal to $\alpha$ ($p \le \alpha$)
    - Fail to reject $H_0$ if p-value is greater than $\alpha$ ($p > \alpha$)


## \alert{Null Hypothesis Significance Testing}: frequent questions

Here is a more typical null distribution and test statistic:

```{r fig.width=3, fig.height=2, fig.align='center'}
ybar <- 0
se <- 1

p <- ggplot(data.frame(x = ybar + 3.75 * c(-se, se)), aes(x)) +
    theme_cowplot() +
    xlab(expression("Test stat if"~H[0]~"is true")) +
    ylab("density") + 
    # ylim(c(0, 0.45)) +
    stat_function(fun = dnorm, 
                  args = list(mean = ybar, sd = se),
                  xlim = ybar + 2.2 * c(-se, se),
                  geom = "area", fill = "blue", alpha = 0.5) +
    stat_function(fun = dnorm, 
                  args = list(mean = ybar, sd = se),
                  xlim = ybar + c(-3.75 * se, -2.2 *se),
                  geom = "area", fill = "orange") +
    stat_function(fun = dnorm, 
                  args = list(mean = ybar, sd = se),
                  xlim = ybar + c(2.2 * se, 3.75 *se),
                  geom = "area", fill = "orange") +
    stat_function(fun = dnorm, args = list(mean = ybar, sd = se)) +
    geom_vline(xintercept = ybar + 2.2 * se, size = 1) +
    annotate("label", 4, 0.45, label = "Observed test stat = 2.2", hjust = "right")

p
```

## \alert{Null Hypothesis Significance Testing}: frequent questions {.t}

::: columns

:::: column

```{r fig.width=3, fig.height=2, fig.align='center'}
p
```

The \alert{p-value} is the sum of the probabilities in the two \textcolor{orange}{orange} regions

::::

:::: column

**Question:** If the test statistic = -2.2, why is the p-value the sum of both \textcolor{orange}{orange} regions"

- Because $H_0$ is no difference, and $H_A$ is *any* difference, not distinguishing between positive or negative differences
- This type of alternative hypothesis is called a "two tailed" test

::::

:::

## \alert{Null Hypothesis Significance Testing}: frequent questions {.t}

::: columns

:::: column

```{r fig.width=3, fig.height=2, fig.align='center'}
p
```

The \alert{p-value} is the sum of the probabilities in the two \textcolor{orange}{orange} regions

::::

:::: column

**Question:** Why is the p-value the probability of "the test statistic or a more extreme value"?

::::

:::


## \alert{Null Hypothesis Significance Testing}: frequent questions {.t}

::: columns

:::: column

```{r fig.width=3, fig.height=2, fig.align='center'}
p
```

The \alert{p-value} is the sum of the probabilities in the two \textcolor{orange}{orange} regions

::::

:::: column

**Question:** Why is the p-value the probability of "the test statistic or a more extreme value"?

- Because if the null distribution is \alert{continuous}, the probability of the exact test statistic will be 0!

::::

:::

## \alert{Null Hypothesis Significance Testing}: frequent questions {.t}

::: columns

:::: column

```{r fig.width=3, fig.height=2, fig.align='center'}
p
```

The \alert{p-value} is the sum of the probabilities in the two \textcolor{orange}{orange} regions

::::

:::: column

**Question:** If we reject $H_0$ at significance level $\alpha = 0.05$? Does that mean our alternative hypothesis is true?

::::

:::

## \alert{Null Hypothesis Significance Testing}: frequent questions {.t}

::: columns

:::: column

```{r fig.width=3, fig.height=2, fig.align='center'}
p
```

The \alert{p-value} is the sum of the probabilities in the two \textcolor{orange}{orange} regions

::::

:::: column

**Question:** If we reject $H_0$ at significance level $\alpha = 0.05$? Does that mean our alternative hypothesis is true?

- NO! Just that we reject the null at $\alpha = 0.05$
- Also we might have **incorrectly** rejected the null!

::::

:::

## Errors in hypothesis testing

There are two types of errors in hypothesis testing

1. Rejecting the null hypothesis when it's true (\alert{false positive})

2. Not rejecting the null hypothesis when it's false (\alert{false negative})

## \alert{False positive}

False positive: Rejecting a true null

* This happens with probability $\alpha$
* Pr[False Positive] is independent of sample size
* Called a “Type I error”

You can decrease your false positive rate by decreasing $\alpha$

## \alert{False negative}

False negative: Failing to reject a false null

* Pr[False Negative] depends on sample size, effect size, and precision
* Also called a “Type II error”

Pr[False Negative] = 1 - statistical power

You can decrease your false negative rate by increasing $\alpha$

You can increase statistical power by increasing your sample size and measurement precision



## Errors

 Decision         | $H_0$ True  | $H_0$ False
------------------|----------|----------
 Reject $H_0$        | Type I   | no error
 Do not reject $H_0$ | no error | Type II
 
 lower $\alpha$ reduces Type I error rate, but raises Type II error rate
 
 both are errors! many methods today choose $\alpha$ to minimize all error, but this is complex.

## Effects of sample size

If $H_0$ is true, sample size has no effect

If $H_0$ is false, statistical power increases with sample size

```{r, fig.align="center", out.width="70%"}
library(pwr)
power <- pwr.t.test(n = 2:512, d = 0.5, sig.level = 0.05, power = NULL, 
                    type = "two.sample", alternative = "two.sided")

df <- tibble(n = rep(2:512, 2), power = c(rep(0.05, 511), power$power),
             group = rep(c("H0 True", "H0 False"), each = 511)) |>
    mutate(group = factor(group, 
                          levels = c("H0 True", "H0 False"), 
                          labels = c(expression(H[0]~True), expression(H[0]~False)))
           )

ggplot(df, aes(n, power, color = group)) +
    facet_grid(. ~ group, labeller = label_parsed) + 
    geom_line(show.legend = FALSE, size = 4) +
    scale_x_log10(breaks = c(8, 64, 512)) +
    scale_y_continuous(limits = c(0, 1)) +
    # scale_color_viridis_d() +
    xlab("Sample size") +
    ylab(expression(
        paste("Probability of rejecting ", italic(H)[0], " (", alpha == 0.05, ")"))) +
    theme_cowplot() +
    theme(
        axis.text = element_text(size = 18),
        axis.title = element_text(size = 24),
        strip.text = element_text(size = 24)
    )

```


