---
title: "BIOL 220 Problem Set 07: Hypothesis tests with a chi-squared null distribution"
subtitle: "Answer Key"
format:
  pdf:
    toc: false
    number-sections: false
    colorlinks: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = TRUE, message = FALSE, 
                      warning = FALSE)

library(cowplot)
library(dplyr)
library(ggplot2)
library(DT)
```


**Due Thursday, March 14, 2024 before lecture**

**Submit your answers via Google Classroom**

## Working with the $\chi^2$ goodness of fit test 

Suppose you have data on the frequencies of three different groups (you can imagine your favorite kinds of things---three fish species, three types of candy, three health behaviors). You would like to know if the the groups and equally represented in your data or not.  Imagine this is what your data look like:

|  group A|  group B|  group C|
|--:|--:|--:|
| 12| 11| 13|

::: {.callout-tip}
## Use the above table to answer questions 1--3.

1. Under the null hypothesis of equal frequencies across groups, what are the expected frequencies for groups A, B, and C? [1 point]

::: {.callout-note appearance="simple"}
## Answer

`12, 12, 12`

:::

2. Calculate the $\chi^2$ statistic for these data [1 point]

::: {.callout-note appearance="simple"}
## Answer

```{r echo = TRUE}
obs <- c(12, 11, 13)
expected <- c(12, 12, 12)

chi2 <- sum(((obs - expected)^2) / expected)
chi2
```

:::

3. What are the degrees of freedom for these data? [1 point]

::: {.callout-note appearance="simple"}
## Answer

`2`

:::

4. With the $\chi^2$ statistic and degrees of freedom you just figured out, use `pchisq` to calculate the $P$-value of the null hypothesis [1 point]

::: {.callout-note appearance="simple"}
## Answer

```{r echo = TRUE}
pchisq(chi2, 2, lower.tail = FALSE)
```

:::

5. Do we reject the null hypothesis? [1 point]

::: {.callout-note appearance="simple"}
## Answer

We fail to reject the null at $\alpha = 0.05$ because the $P$-value is greater than 0.05

:::


:::



# Diabetes prevelance in Hawaiʻi 

[Grandinetti et al. (2007)](https://pubmed.ncbi.nlm.nih.gov/17682354/) studied prevalence of diabetes in North Kōhala on the Island of Hawaiʻi. They found significant disparities in diabetes prevalence across ethnic groups, with Native Hawaiian and Asian individuals at higher risk of diabetes. This increased risk could not be explained by shared lifestyle risk factors, indicating that other, unmeasured, drivers were likely at play (side note from Andy: perhaps colonialism).

The table below summarizes the results of Grandinetti et al. (2007) with respect to diabetes prevelance by ethnic group.


```{r}
x <- matrix(c("Caucasian", 13, 295,
              "Filipino", 36, 186,
              "Hawaiian", 100, 526,
              "Japanese", 40, 190,
              "Other/mixed non-Hawaiian", 39, 255), 
            ncol = 3, byrow = TRUE)
y <- matrix(as.numeric(x[, -1]), ncol = 2)
y[, 2] <- y[, 2] - y[, 1]
y <- t(y)

colnames(y) <- x[, 1]
rownames(y) <- c("Yes diabetes", "No diabetes")

knitr::kable(y)
```

::: {.callout-tip}
## Use the above table to answer questions 6--9.

6. To address the question of "are there differences in diabetes prevalence across ethnic groups?" what should our null hypothesis be? [1 point]

::: {.callout-note appearance="simple"}
## Answer

$H_0$: The per capita proportion of diabetes is the same across all ethnic groups 

:::

7. If you were to use a $\chi^2$ distribution to represent this null hypothesis, what are the correct degrees of freedom? [1 point]

    - 1
    - 2
    - 4
    - 5
    - 10

::: {.callout-note appearance="simple"}
## Answer

(Number of rows - 1) $\times$ (Number of columns - 1) $= 1 \times 4 = 4$ 

:::

8. Use R code to make a matrix of the diabetes data. To answer this question, paste your R code into the google form [1 point]

::: {.callout-note appearance="simple"}
## Answer

```{r echo = TRUE}
# as long as your R code makes the correct matrix, thatʻs ok
# correct code will probably looks something like

diabetes <- matrix(c(13, 282, 36, 150, 100, 426, 40, 150, 39, 216), 
                   nrow = 2)
diabetes

```

:::

9. Use `chisq.test` to conduct a $\chi^2$ null hypothesis test. Report the test statistic and $P$-value in the google form [1 point]

::: {.callout-note appearance="simple"}
## Answer

```{r echo = TRUE}
chisq.test(diabetes)

```

:::

:::
  
## $\chi^2$ distribution

::: {.callout-tip}
## $\chi^2$ distribution

10. Which of the following figures correctly shows a $\chi^2$ sampling distribution with 6 degrees of freedom and a test statistic (vertical line) consistent with a $P$-value of 0.04. [1 point]

```{r, fig.align="center", out.width="75%"}


df1 <- tibble(x = seq(0, 20, 0.01), y = dchisq(x, 6))
df2 <- tibble(x = seq(0, 20, 0.01), y = dchisq(x, 2))
ts1 <- qchisq(0.04, 6, lower.tail = FALSE)
ts2 <- qchisq(0.04, 6, lower.tail = TRUE )
gp1 <- ggplot(df1, aes(x, y)) +
  geom_area(color = "black") +
  xlab(expression(chi^2)) +
  ylab("Density") +
  theme_cowplot()

gp2 <- gp1 %+% df2 + geom_vline(xintercept = ts1, color = "grey", linetype = "dashed", size = 2)
gp3 <- gp1 + geom_vline(xintercept = ts2, color = "grey", linetype = "dashed", size = 2)
gp4 <- gp1 %+% df2 + geom_vline(xintercept = ts2, color = "grey", linetype = "dashed", size = 2)
gp1 <- gp1 + geom_vline(xintercept = ts1, color = "grey", linetype = "dashed", size = 2)

plot_grid(gp3, gp1, gp4, gp2, nrow = 2, labels = "AUTO")

```

::: {.callout-note appearance="simple"}
## Answer

**B**

:::

:::
