---
title: "BIOL 220: Analysis of Variance"
output: binb::metropolis
header-includes:
  - \usepackage{array}
  - \usepackage{xcolor}
  - \usepackage{colortbl}
  - \usepackage[export]{adjustbox}
  - \usepackage{pdfpages}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, 
                      message = FALSE, eval = TRUE)

library(cowplot)
library(ggplot2)
library(dplyr)
library(viridis)
library(glue)
library(stringr)
library(tidyr)
library(googlesheets4)
library(extraDistr)
library(purrr)
library(magrittr)
```



## Learning Objectives


1. Know when to use analysis of variance 
2. Understand that the null hypothesis for analysis of variance is no difference in means across groups or values of explanatory variables
3. Understand how to decompose variance into its sources
4. Understand the problem of multiple hypothesis testing


# To the board...




## Example ANOVA for multiple groups in *R*


\small

```{r echo = TRUE}
# make some pretend data
groups <- rep(LETTERS[1:3], each = 10)
y <- 1.2 * as.numeric(as.factor(groups)) + rnorm(30, mean = 0, sd = 0.5)

dat_groups <- data.frame(group = groups, y = y)

head(dat_groups, n = 3)
```


## Example ANOVA for multiple groups in *R*


\small

```{r echo = TRUE, fig.align='center', fig.width=3, fig.height=3}
ggplot(dat_groups, aes(x = group, y = y)) + geom_boxplot()
```


## Example ANOVA for multiple groups in *R*


\small

```{r echo = TRUE, fig.align='center', fig.width=3, fig.height=2}
ggplot(dat_groups, aes(x = group, y = y)) + 
    geom_jitter(aes(color = group))
```

## Example ANOVA for multiple groups in *R*


\small

```{r echo = TRUE}
mod_group <- lm(y ~ group, data = dat_groups)
anova(mod_group)
```


## Assumptions of ANOVA for groups

- Random and independent samples
- Normal distribution for each group
- Equal variances within each group

## What to do if assumptions are not met?

- Do ANOVA anyway (if deviations are minor)
- Transform
- More advanced stats

## Interpreting ANOVA for groups

ANOVA does not tell you which group(s) is/are different!  Just that a difference exists

\small

```{r echo = FALSE, fig.align='center', fig.width=3, fig.height=3}
ggplot(dat_groups, aes(x = group, y = y)) + geom_boxplot()
```

## Interpreting ANOVA for groups

If we want to know which groups are different, we need to do a "post hoc" test

- Post hoc tests are designed to account for multiple comparisons
- What are multiple comparisons and why can't we just do multiple $t$-tests?

## Pos hoc tests and multiple comparisons

Which groups are different?  Just do $t$-tests between all pairwise comparisons, right?

- The problem is that doing multiple $t$-tests leads to a false positive rate $> \alpha$
- If we do 3 pairwise comparisons with $t$-tests and $\alpha = 0.05$ for each one, then probability of at least 1 false positive is:

```{r echo = TRUE}
prob_false_positive <- 1 - dbinom(0, size = 3, prob = 0.05)
prob_false_positive
```


## Pos hoc tests and multiple comparisons

```{r echo = TRUE}
prob_false_positive <- 1 - dbinom(0, size = 3, prob = 0.05)
prob_false_positive
```

Each $t$-test is like a coin toss with "heads = false positive", the more you do, the more likely you'll get at least one "heads = false positive"


## Pos hoc tests and multiple comparisons

So we use a post hoc test:

\small

```{r echo = TRUE}
# annoying, but we have to pass the model through the `aov` function
aov_group <- aov(mod_group) 
```

```{r echo = TRUE, eval = FALSE}
# "Tukey Honest Significant Difference"
TukeyHSD(aov_group)
```


## Pos hoc tests and multiple comparisons

So we use a post hoc test:

\small
```{r echo = TRUE}
# "Tukey Honest Significant Difference"
TukeyHSD(aov_group)
```


## Pos hoc tests and multiple comparisons

The adjusted Confidence Intervals and $P$-values make sure that the overall $\alpha$ remains at the desired level, usually 0.05

