---
title: "BIOL 220: Analysis of Variance"
output: binb::metropolis
header-includes:
  - \usepackage{array}
  - \usepackage{xcolor}
  - \usepackage{colortbl}
  - \usepackage[export]{adjustbox}
  - \usepackage{pdfpages}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, 
                      message = FALSE, eval = TRUE)

library(cowplot)
library(ggplot2)
library(dplyr)
library(viridis)
library(glue)
library(stringr)
library(tidyr)
library(googlesheets4)
library(extraDistr)
library(purrr)
library(magrittr)
```


# Multi-factor ANOVA 

## Learning Objectives


1. What is a multi-factor ANOVA, when do you use it, and how?


# To the board...


## Example multi-factor ANOVA in *R*


\small

```{r echo = TRUE}
# make some pretend data
restoration <- factor(rep(c("restored", "control"), each = 30))
impact <- factor(rep(c("low", "high"), 30), 
                 levels = c("low", "high"))

# compute expected response
fish_biomass <- -1 * as.numeric(impact) + 3 * as.numeric(restoration) -
  1 * as.numeric(impact) * as.numeric(restoration)

# add some noise
set.seed(123)
fish_biomass <- fish_biomass + rnorm(60, mean = 0, sd = 1)

dat <- data.frame(restoration, impact, fish_biomass)
```


## Example multi-factor ANOVA in *R*


\small

```{r}
# have a look at pretend data
head(dat)
```


## Example multi-factor ANOVA in *R*


\small

```{r echo = TRUE, eval = FALSE}
ggplot(dat, aes(x = restoration, y = fish_biomass, 
                fill = impact)) + 
  geom_boxplot() + 
  scale_fill_brewer(palette = "YlGnBu", direction = -1)
```


## Example multi-factor ANOVA in *R*


```{r echo = FALSE, fig.align='center', fig.width=3, fig.height=2}
gp <- ggplot(dat, aes(x = restoration, y = fish_biomass, 
                fill = impact)) + 
  geom_boxplot() + 
  scale_fill_brewer(palette = "YlGnBu", direction = -1)
gp
```

## Example multi-factor ANOVA in *R*


\small

```{r echo = TRUE}
mod <- lm(fish_biomass ~ impact + restoration, data = dat)
anova(mod)
```

## Example multi-factor ANOVA in *R*

How do we see the estimate of the effects?
\footnotesize

```{r echo = TRUE}
summary(mod)
```

## Example multi-factor ANOVA in *R*


But looking at the graph, it also seems like the effect of restoration depends on human impact

```{r fig.align='center', fig.width=3, fig.height=2}
gp
```

The two effects \alert{interact}

## Example multi-factor ANOVA in *R* **with interactions**

The asterisk `*` between terms allows for interactions

\small

```{r echo = TRUE}
mod_interactions <- lm(fish_biomass ~ impact * restoration, 
                       data = dat)
anova(mod_interactions)
```


## Example multi-factor ANOVA in *R* **with interactions**

And the interaction effect is meaningful!

\footnotesize

```{r echo = TRUE}
summary(mod_interactions)
```


Higher human impact *reduced* the effect of restoration


## Example multi-factor ANOVA in *R* **with interactions**

If you want to do a post-hock test, how would you do it?

## Example multi-factor ANOVA in *R* **with interactions**

If you want to do a post-hock test, how would you do it? *Just like before!*

\small

```{r echo = TRUE, eval = FALSE}
aov_interactions <- aov(mod_interactions)
TukeyHSD(aov_interactions)
```

\tiny

```{r}
aov_interactions <- aov(mod_interactions)
TukeyHSD(aov_interactions)
```

## Example multi-factor ANOVA in *R* **with interactions**


\tiny

```{r}
foo <- TukeyHSD(aov_interactions)
foo[3]
```


