---
title: "BIOL 220 Graphics"
output: binb::metropolis
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
library(cowplot)
library(ggplot2)
library(scales)
library(dplyr)
library(tidyr)

x <- read.csv("../../data/Focal_Bird_Morphology.csv")
xname <- data.frame(species = tolower(c("JAWE", "RBLE", 
                                        "RVBU", "RWBU",
                                        "WRSH")), 
                    full_name = c("Japanese white-eye", "red-billed leiothrix", 
                                  "red-vented bulbul", "red-whiskered bulbul", 
                                  "white-rumped shama"))

x <- full_join(x, xname)
```

## Ka hua \`ōlelo o ka ha\`awina

\centering \Large
\alert{kuhikuhi}


## Learning Objectives

1. Distinguish between \alert{exploratory} and \alert{explanatory} figures
2. Identify what makes a good graph
3. Identify best practices in figure design
4. Identify misleading graphics and fix them
5. Understand how variable types determine figure design 

# 5. Understand how variable types determine figure design 

## 5. Understand how variable types determine figure design 

We make use of the *R* package `ggplot2` for figure design

The basic building blocks are:

```{r eval = FALSE, echo=TRUE}
library(ggplot2)

ggplot(data = "the data.frame we're plotting", 
       mapping = aes(x = "column with data for x-axis", 
                     y = "column with data for y-axis")) +
    geom_"type of plot"
```


## 5. Understand how variable types determine figure design 

::: columns

:::: {.column width=0.2}

**Explanatory:** Numerical 
\vspace{1em}

**Response:** Numerical
\vspace{1em}

**Figure:** Scatter plot, `geom_point`

::::

:::: {.column width=0.8}


```{r fig.align='center', fig.width=3, fig.height=3}
set.seed(200)
dat <- data.frame(treatment = rep(LETTERS[1:3], c(12, 30, 18)))

dat$type <- as.factor(dat$treatment) |> 
    as.numeric() |> 
    rnorm(nrow(dat), mean = _) |> 
    cut(breaks = 4) |> 
    as.numeric() |> 
    (\(x) paste0("type_0", x))()

dat$x <- runif(nrow(dat), 0, 20)
dat$y <- 0.5 * dat$x + 4 * as.numeric(as.factor(dat$treatment)) + rnorm(nrow(dat))


ggplot(dat, aes(x = x, y = y)) +
    geom_point()
```

::::

:::

## 5. Understand how variable types determine figure design 

::: columns

:::: {.column width=0.2}

**Explanatory:** Categorical *or* Ordinal  
\vspace{1em}

**Response:** Numerical
\vspace{1em}

**Figure:** Jitter plot, `geom_jitter`

::::

:::: {.column width=0.8}

```{r fig.align='center', fig.width=4, fig.height=4}
ggplot(dat, aes(x = treatment, y = y, color = treatment)) +
    geom_jitter() +
    scale_color_viridis_d() +
    theme(legend.position = "none")
```

::::

:::

## 5. Understand how variable types determine figure design 

::: columns

:::: {.column width=0.2}

**Explanatory:** Categorical *or* Ordinal  
\vspace{1em}

**Response:** Numerical
\vspace{1em}

**Figure:** Box plot, `geom_boxplot`

::::

:::: {.column width=0.8}

```{r fig.align='center', fig.width=4, fig.height=4}
ggplot(dat, aes(x = treatment, y = y, fill = treatment)) +
    geom_boxplot() +
    scale_fill_viridis_d() +
    theme(legend.position = "none")

```

::::

:::


## 5. Understand how variable types determine figure design 

::: columns

:::: {.column width=0.2}

**Explanatory:** Categorical *or* Ordinal  
\vspace{1em}

**Response:** Categorical *or* Ordinal 
\vspace{1em}

**Figure:** Stacked bar plot, `geom_bar`

::::

:::: {.column width=0.8}

```{r fig.align='center', fig.width=4, fig.height=4}
ggplot(dat, aes(x = treatment, fill = type)) +
    geom_bar() +
    scale_fill_viridis_d(option = "inferno")

```

::::

:::

## 5. Understand how variable types determine figure design 

::: columns

:::: {.column width=0.2}

**Explanatory:** Categorical *or* Ordinal  
\vspace{1em}

**Response:** Categorical *or* Ordinal 
\vspace{1em}

**Figure:** Stacked bar plot (scaled to percent), `geom_bar(position = "fill")`

::::

:::: {.column width=0.8}

```{r fig.align='center', fig.width=4, fig.height=4}
ggplot(dat, aes(x = treatment, fill = type)) +
    geom_bar(position = "fill") +
    scale_fill_viridis_d(option = "inferno") +
    ylab("Percent")
```

::::

:::

## 5. Understand how variable types determine figure design 

::: columns

:::: {.column width=0.2}

**Explanatory:** "none" (i.e. you just want to check out response)
\vspace{1em}

**Response:** Numerical
\vspace{1em}

**Figure:** Histogram, `geom_histogram`

::::

:::: {.column width=0.8}

```{r fig.align='center', fig.width=4, fig.height=4}
ggplot(dat, aes(x = y)) +
    geom_histogram(color = "black", binwidth = 4)

```

::::

:::

## 5. Understand how variable types determine figure design 

::: columns

:::: {.column width=0.2}

**Explanatory:** "none" (i.e. you just want to check out response)
\vspace{1em}

**Response:** Categorical *or* Ordinal 
\vspace{1em}

**Figure:** Bar plot, `geom_bar`

::::

:::: {.column width=0.8}

```{r fig.align='center', fig.width=4, fig.height=4}
ggplot(dat, aes(x = type)) +
    geom_bar()

```


::::

:::

## 1. Distinguish between \alert{exploratory} and \alert{explanatory} figures

For \alert{exploratory plots}: Don't worry about pretty colors, labels, names, etc. The goal is for you to understand the data.

For \alert{explanatory plots}: Focus on all of this. You must consider your audience may, for example, consist of:

- People unfamiliar with your data          
- People with different vision, and/or black-and-white printing

## 1. Distinguish between \alert{exploratory} and \alert{explanatory} figures

Let's use an example 

```{r, echo = TRUE}
# Data on morphology of some non-native birds in Hawaiʻi
# from Gleditsch & Sperry (2020) *Ecology and Evolution*
birds <- read.csv("../../data/bird_morpho.csv")
head(birds)
```


## 1. Distinguish between \alert{exploratory} and \alert{explanatory} figures

```{r}
gp1 <- ggplot(birds, aes(x = mass, y = wing)) +
    geom_point() 


gp2 <- ggplot(birds, aes(x = mass, y = wing)) +
    geom_point(size = 2, alpha = 0.5) +
    xlab("Body mass (g)") +
    ylab("Wing length (mm)") +
    geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
    theme_cowplot() +
    theme(axis.text = element_text(size = 18),
          axis.title = element_text(size = 20))
```

::: columns

:::: column

\Large \alert{Exploratory}

```{r, fig.asp = 1, fig.width = 4}
gp1
```

::::

:::: column

\Large \alert{Explanatory}
```{r, fig.asp = 1, fig.width = 4}
gp2
```

::::

:::

## 1. Distinguish between \alert{exploratory} and \alert{explanatory} figures

Contrast code using **ggplot2** (we fully explain in Lab 02)

::: columns

:::: column

\Large \alert{Exploratory}

\tiny
```{r, echo=TRUE, eval=FALSE}
# minimal code needed
ggplot(birds, aes(x = mass, y = wing)) +
    geom_point() 
```

::::

:::: column

\Large \alert{Explanatory}

\tiny
```{r, echo=TRUE, eval=FALSE}
# plenty more code needed
ggplot(birds, aes(x = mass, y = wing)) +
    geom_point(size = 2, alpha = 0.5) +
    xlab("Body mass (g)") +
    ylab("Wing length (mm)") +
    geom_smooth(method = "lm", formula = y ~ x, 
                se = FALSE) +
    theme_cowplot() +
    theme(axis.text = element_text(size = 18),
          axis.title = element_text(size = 20))
```

::::

:::



# Keys to Good Data Visualization

## Four keys of good data visualization

1. Show the data
2. Make patterns in the data easy to see
3. Represent magnitudes honestly
4. Display graphical elements clearly

## Mistake 1: Hide the data

How to hide data:

- Provide only statistical summaries
- Over-plotting

How to show data:

- Present all of the data points, while allowing all to be seen

## Mistake 1: Hide the data

This plot hides the variation within species

```{r fig.align='center', fig.width=4, fig.height=3}
bd <- birds[birds$species %in% c("rwbu", "wrsh", "jawe"), ]

ggplot(bd, aes(x = full_name, y = mass, fill = full_name)) +
    geom_bar(stat = "summary", fun = "mean") +
    theme(legend.position = "none") +
    xlab("") + ylab("Mass (g)") +
    scale_fill_viridis_d() +
    scale_x_discrete(labels = function(x) {
        strwrap(x, width = 10, simplify = FALSE) |> 
            sapply(FUN = paste, collapse = "\n")
    })
```


## Mistake 1: Hide the data

This plot hides the density of observations

```{r fig.align='center', fig.width=4, fig.height=3}
ggplot(bd, aes(x = full_name, y = mass, color = species)) +
    geom_point() +
    theme(legend.position = "none") +
    xlab("") + ylab("Mass (g)") +
    scale_color_viridis_d() +
    scale_x_discrete(labels = function(x) {
        strwrap(x, width = 10, simplify = FALSE) |> 
            sapply(FUN = paste, collapse = "\n")
    })
```

## Mistake 1: Hide the data

**Fixed!** This plot shows all of the observations

```{r fig.align='center', fig.width=4, fig.height=3}
ggplot(bd, aes(x = full_name, y = mass, color = species)) +
    geom_jitter(alpha = 0.5) +
    scale_color_viridis_d() +
    theme(legend.position = "none") +
    xlab("") + ylab("Mass (g)") +
    scale_x_discrete(labels = function(x) {
        strwrap(x, width = 10, simplify = FALSE) |> 
            sapply(FUN = paste, collapse = "\n")
    })
```


## Mistake 2: Make patterns hard to see

How to hide patterns:

- Make one plot and call it good
- Use unreasonable scales
- Arrange factors nonsensically

How to make patterns in the data easy to see:

- Choose the best from multiple potential plots
- Use appropriate scales
- Arrange factors meaningfully


## Mistake 2: Make patterns hard to see

Non-intuitive ordering of categories hides patterns

```{r fig.align='center', fig.width=4, fig.height=3}
d <- data.frame(Treatment = rep(c("control", "low dose", "high dose"), 
                            each = 20))

set.seed(123)

d$Response <- rnorm(nrow(d))

d$Response[d$Treatment == "control"] <- 
    d$Response[d$Treatment == "control"] + 3
d$Response[d$Treatment == "low dose"] <- 
    d$Response[d$Treatment == "low dose"] + 8

d$Response[d$Treatment == "high dose"] <- 
    d$Response[d$Treatment == "high dose"] + 10

pal <- hsv(c(0.08, 0.12, 0.8), 
           c(1.0, 0.7, 1.0), 
           c(0.8, 0.9, 0.4))
ggplot(d, aes(Treatment, Response, color = Treatment)) +
    geom_jitter() +
    scale_color_manual(values = pal[c(3, 1, 2)])
```


## Mistake 2: Make patterns hard to see

*Fixed!* Intuitive ordering of categories emphasizes patterns

```{r fig.align='center', fig.width=4, fig.height=3}
d$Treatment <- factor(d$Treatment, 
                      levels = c("control", "low dose", "high dose"))
gtrial <- ggplot(d, aes(Treatment, Response, color = Treatment)) +
    geom_jitter() +
    scale_color_manual(values = pal[c(3, 1, 2)])

gtrial
```


## Mistake 2: Make patterns hard to see

Poorly chosen scale hides patterns

```{r fig.align='center', fig.width=4, fig.height=3}
gtrial + scale_y_continuous(limits = c(0, 400))
```


## Mistake 3: Display Magnitudes Dishonestly

How to represent magnitudes *dis*honestly:

- Starting/stopping axes at arbitrary values

How to represent magnitudes *honestly*:

- Don't do that!


## Mistake 3: Display Magnitudes Dishonestly

We've already seen this once

```{r fig.align='center', fig.width=4, fig.height=3}
gtrial + scale_y_continuous(limits = c(0, 400))
```


## Mistake 3: Display Magnitudes Dishonestly

::: columns

:::: column
\includegraphics[width = \textwidth]{../../img/ws_magnitude-honest.pdf}
::::

:::: column
**Note:** there are a small number of cases where barplots can *honestly* start at a value other than zero
::::

:::

## Mistake 4: Display graphical elements unclearly

How to display graphical elements unclearly:

- Accept default plotting options
- Do not consider how a diverse audience will interpret your figure

How to display graphical elements clearly:

- Reflect on plot design
    - Are labels big enough?
    - Can all elements be clearly seen?
    - Is sizing and spacing appropriate?
- Consider diverse audiences and how they will interact with your plot


## Mistake 4: Display graphical elements unclearly

Can you read the x-axis?

```{r fig.width=4.5, fig.height=3}
ggplot(birds, aes(x = full_name)) +
    geom_bar() +
    xlab("") +
    ylab("Sample size")
```

## Mistake 4: Display graphical elements unclearly

Flipping improves readability

```{r fig.width=4.5, fig.height=3}
ggplot(birds, aes(y = full_name)) +
    geom_bar() +
    xlab("Sample size") +
    ylab("")
```


## Mistake 4: Display graphical elements unclearly

Flipping improves readability...and we should probably re-order

```{r fig.width=4.5, fig.height=3}
birds$full_name <- factor(birds$full_name, 
                          levels = names(sort(table(birds$full_name))))
ggplot(birds, aes(y = full_name)) +
    geom_bar() +
    xlab("Sample size") +
    ylab("")
```


## Mistake 4: Display graphical elements unclearly

Consider colorblindness


```{r fig.width=9, fig.height=4}
gp3 <- ggplot(bd, aes(x = mass, y = wing, color = species)) +
    geom_point() + 
    xlab("Body mass (g)") +
    ylab("Wing length (mm)") +
    theme_cowplot() +
    theme(axis.text = element_text(size = 18),
          axis.title = element_text(size = 20), 
          legend.position = c(0.6, 0.2))


plot_grid(
    gp3 + ggtitle("no color blindness"), 
    gp3 + scale_color_manual(values = c("#A59A71", "#908048", "#768FF5")) +
        ggtitle("red-green"), 
    gp3 + scale_color_manual(values = c("#ED7377", "#51B293", "#60ADB7")) +
        ggtitle("blue-yellow"), 
    nrow = 1
)
```

