---
title: "BIOL 220: Regression"
output: binb::metropolis
header-includes:
  - \usepackage{array}
  - \usepackage{xcolor}
  - \usepackage{colortbl}
  - \usepackage[export]{adjustbox}
  - \usepackage{pdfpages}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, 
                      message = FALSE, eval = TRUE)

library(cowplot)
library(ggplot2)
library(dplyr)
library(viridis)
library(glue)
library(stringr)
library(tidyr)
library(googlesheets4)
library(extraDistr)
library(purrr)
library(magrittr)
```





## Showing steps for calculating regression models

Make some fake tooth decay data

```{r}
# this makes all "random" numbers the same every time
# we run this code
set.seed(12) 

# random dates
deltaC14 <- runif(18, 0, 600)

# simulate date of birth from slope, intercept, 
# and residual error
dateOfBirth <- 1992.27 - 0.05326 * deltaC14 + 
    rnorm(18, sd = 10.99)

```

## Showing steps for calculating regression models

Make some fake tooth decay data

```{r}
# combine into data.frame
tooth.data <- data.frame(deltaC14 = deltaC14, 
                         dateOfBirth = dateOfBirth)

# have a look
head(tooth.data)

```



## Showing steps for calculating regression models

Plot it

```{r fig.width=3.5, fig.height=3, fig.align='center', out.width="50%"}
ggplot(tooth.data, aes(deltaC14, dateOfBirth)) +
    geom_point()
```


## Showing steps for calculating regression models

Make a linear model

```{r}
tooth.model <- lm(dateOfBirth ~ deltaC14, 
                  data = tooth.data)

```


## Showing steps for calculating regression models

Get info on null hypothesis tests

```{r eval=FALSE}
summary(tooth.model)

```

\tiny

```{r echo=FALSE}
summary(tooth.model)
```

Notice that `F-statistics` stuff, we'll get there! But not yet


## Showing steps for calculating regression models

Get confidence intervals of coefficient estimates

```{r}

confint(tooth.model)

```